A Middleware for Symmetric Peer-to-Peer Applications

### 2023-03-17 tpds-23/tpds-23.tex
...

Real-time networked applications allow multiple users to interact remotely and
yet share the same experience.
Examples of these \emph{symmetric distributed applications}~\cite{gals} are
collaborative documents, watch parties, and multi-player games.
%~\cite{TODO:collab}.
%
In order to reproduce the exact behavior in multiple devices, the application
must be able to synchronize time and execution across the network.
In addition, since users can interact asynchronously with the application,
event occurrences must somehow be synchronized back across devices.
%
This problem relates to consistency in distributed systems, and is also known
as \emph{state melding}~\cite{melding} in the context of virtual worlds, which
adds the requirement of a time deadline to reach consistency.

@@

In this work, we propose a middleware for \emph{symmetric peer-to-peer
applications}, which do not rely on central servers for coordination.
Peers in the application form a dynamic network graph and communicate only
with direct neighbours, as in typical unstructured peer-to-peer
networks~\cite{p2p.survey}.
%
Events are flooded in the graph and are applied with a small deadline delay to
compensate the network latency.
To deal with events received out of order or too late, a time machine can
rollback a peer to a previous state and reapply the events in order, such that
a single consistent timeline is established across the network.
%
Our main contribution is the design of a middleware to ensure that all peers
    (a) meet event deadlines,
    (b) advance in sync in real time, and
    (c) can leave and join the network and remain symmetric.
%
Our middleware provides \emph{perceptive consistency}~\cite{melding}, which
ensures that an event triggered by any peer at time $t$ is perceived by all
peers simultaneously at time $t+\Delta$.
%
As the main limitation, the middleware cannot ensure \emph{instantaneity} for
events in most scenarios, which would require $\Delta$ to be less than a
user-observable threshold~\cite{melding}.
We also that assume peers are non malicious in the sense that they do not lie
about the time in which events occurred locally.

We perform simulations with over 20 peers in a 5-hop mixed topology with cycles
and straight lines.
We vary the network latency and peer churn, as well as the frequency and
deadlines of events.
%
We then measure (a) the recurrence of time travels, (b) the real-time pace of
peers, and (c) the recoverability from churns, such that they validate our
goals above, respectively.
%
As results, we show that goals (b) and (c) are met even under extreme
conditions: peers exhibit a time mismatch of under $0.2\%$ and take less than
$1.5s$ to recover from churns.
%
For the goal (a), we analyze the scenarios in which the middleware can meet
event deadlines with the desired performance.
As an example, peers need to roll back only $3\%$ of the time under a network
latency of $50ms$ and 25 events per minute.
%
We consider this overhead negligible for soft real-time applications that do
not depend on instantaneous network feedback, such as chats, watch parties, and
turn-based games.

@@

Croquet~\cite{croquet,croquet2,croquet.site} guarantees bit-identical real-time
behavior for users in collaborative distributed environments.
%
As detailed in Figure~\ref{fig.algos}, a centralized \emph{reflector} issues
fixed periodic ticks (e.g., every $50ms$), such that all nodes in the network
advance together according to a synchronized global shared clock.
If a user generates an event, it is sent back to the reflector, which
broadcasts the event in the next tick, which all nodes apply in sync.
%
This mechanism is similar to \emph{bucket synchronization}~\cite{gautier}, in
which events are assigned to time buckets to be applied later.
%
Croquet takes periodic snapshots of the whole application state in order to
support late joins to a running session.
The new node just needs to request the latest snapshot and simulate the
remaining events to reach the current running state.
%
As indicated in Figure~\ref{fig.table}, Croquet relies on a centralized
network (a), in which nodes advance in sync (b), and in which event broadcasts
and outcomes depend on the central server to be online (c).

@@

As it is the case for the discussed work, we consider that middlewares for
symmetric applications, which provide bit-identical behavior, must comply with
\emph{perceptive consistency}~\cite{melding}, as follows:

@@

An antagonistic approach to deterministic reactions to events is to model the
applications with Conflict-free Replicated Data Types~\cite{crdts}.
CRDTs are designed in such a way that all operations are commutative, so that
the order in which they are applied does not affect the final outcome.
%
As detailed in Figure~\ref{fig.algos}, peers can communicate operations
directly to each other with no central authority.
Also, since operations need not to be ordered, they can be applied at the
very first moment they are generated or received, even if the originating peers
are offline.
%
Note that it is not possible to timestamp commutative operations under a unique
global shared clock.
Hence, there is no notion of a timeline in which peers go through bit-identical
steps, but they do eventually reach the same state~\cite{crdts.eventual}.
%
On the one hand, the main advantage of CRDTs is to support local-first
software~\cite{local}, since they can work offline in the same way as online.
On the other hand, they provide very restrictive APIs (the CRDTs themselves),
and do not support real-time consensus among peers~\cite{crdts.consensus}.
%
As indicated in Figure~\ref{fig.table}, CRDTs work on
    (a) decentralized networks, in which
    (b) peers advance independently, and in which
    (c) operations can be applied immediately, even while offline.
%
Automerge~\cite{p2p.automerge} along with its accompanying
Hypermerge~\cite{p2p.pushpin} peer-to-peer infrastructure is an example of
decentralized CRDTs.

The main advantage of CRDTs is that they can be designed in such a way that
concurrent modifications to shared state can preserve intent~\cite{peritext},
regardless of the order in which they are applied.
%
This is not the case for the symmetric application middlewares described in the
previous section.
Nevertheless, programmers can still use CRDTs on top of a middleware, since
they are nothing more than data structures and merging algorithms.

CRDTs provide \emph{delayed consistency}~\cite{melding}, as follows:

@@

Regarding single-user local time machines, the video game Braid%
\footnote{Braid: \url{http://braid-game.com/}}
is designed on the assumption that players have unrestricted rollbacks as part
of the game mechanics.
Like Fusion, the implementation is based on continuous memory snapshots,
instead of deterministic simulation as we propose in this work.
%
Schuster~\cite{tml.js} proposes a simple API for time travelling in
event-driven JavaScript:
    \code{init()} to return an initial state;
    \code{handle(evt,mem)} to process events and modify state; and
    \code{render(mem)} to output the current state.
As we describe next, we use a similar approach by separating state mutation
from rendering~\cite{tml.alive}.

@@

Our middleware is written in C and relies on the library \emph{SDL}%
\footnote{SDL: \url{http://www.libsdl.org}}
to deal with the network, timers, and input events, such as mouse clicks and
key presses.
The full source code%
\footnote{Code: \url{http://github.com/fsantanna/p2p-tml/}}
is under 500 lines of code.

@@

In Figure~\ref{fig.middleware}, a peer is depicted as an application using an
API to communicate with the middleware.
The application itself is not aware of the network and has no direct access to
other peers, which communicate transparently through the middleware.
The application generates asynchronous events that go through the middleware,
which synchronizes them with a timestamp with a deadline delay ($\Delta$)
scheduled to the future, such that all peers can satisfy (also known as
\emph{local lag}~\cite{mauve}]).
The middleware controls the execution of the application by issuing ticks and
synchronized events.
Note that ticks need not to be broadcast, only $\Delta$ events.

The main job of the middleware is to deal with the uncertainties of the
network in order to provide a consistent timeline and preserve the symmetric
behavior across peers.
As described in the Introduction, the middleware ensures that all peers
    (a) meet event deadlines,
    (b) advance in sync in real time, and
    (c) can leave and join the network and remain symmetric:
%
\begin{enumerate}
\item \textbf{Event Deadlines:}
Due to the inherent latency of networks, a peer (through the middleware)
generates events with a deliberate $\Delta$ so that they can reach other peers
in time to be applied in sync.
%We first assume a predetermined delay for the sake of simplicity.
Nevertheless, a peer ahead of time may receive an event that should have been
applied in the past, which characterizes an optimistic approach to choose
$\Delta$, since it assumes that inconsistencies may occur~\cite{melding}.
%, even considering its $\Delta$.
%
In this case, the middleware automatically rolls back the peer, applies the
event at the correct time, and then fast forwards the peer, re-applying all
remaining events up to the real time.
As we detail further, rolling back a peer requires to simulate the whole
application execution and event occurrences from the beginning.
As an optimization, the middleware takes periodic snapshots locally to avoid
full simulation.
%
\item \textbf{Time Synchronization:}
Since peers run in different machines, their timelines will inevitably be out
of sync because applications are launched locally at different times, and
because internal clocks may diverge over time (e.g., from rollbacks or timer
polling inaccuracies).
The middleware assumes the maximum time among all peers the be the
``correct real time'', so that other peers must fast forward to catch up with
it.
Therefore, in order to determine the maximum time and synchronize the clocks,
the middleware proceeds as follows:
    (a) peers broadcast \code{SYNC} events periodically with their local
        times, and
    (b) if a peer is behind a received \code{SYNC}, it advances its frames
        proportionally faster to catch up in time.
%
To avoid abrupt visual changes, the middleware makes smooth time transitions
that take up to $1s$ to complete.
%
\item \textbf{Peer Churn:}
Considering that we target dynamic peer-to-peer networks, it is important that
the middleware can handle arrival and departure of peers seamlessly.
Note that high churn may lead to intermittent network partitions.
Given our unstructured approach, nothing needs to be done when a peer leaves
the network.
However, when a peer joins the network, the middleware needs to ensure that it
receives from other peer all events that ever happened, in order.
This is trivial since all peers already need to hold the full event history (as
described in item (a)).
% including their absolute timestamps in the shared timeline.
\end{enumerate}

As a limitation of our work, regarding items (a) and (b), we assume that peers
never lie about their local clocks, which would affect the event timestamps,
allowing a peer to modify the global timeline with nonsensical information.
%
For this reason, we assume a permissioned peer-to-peer network with an external
mechanism for authentication.

@@

The middleware expects to take full control of the application execution in
order to manipulate its timeline and disseminate events in the network.
For this reason, the basic programming API is to call a \code{loop} function
passing a set of callbacks, which the middleware uses to communicate with the
application:

@@

Note that using callbacks is a already common technique for event-driven
programming, such as in Games, GUIs, and interactive applications in general,
sharing similarities with \emph{MVC}, \emph{Publish-Subscribe}, and
\emph{Observer} patterns~\cite{meyer,nystrom}.
%
Furthermore, the default run-to-completion semantics of callbacks is key to
ensure the desired determinism, since unlike threads, they run sequentially and
atomically from start to end~\cite{events,threads}.

@@

The middleware calls \code{cb\_sim} every frame or event occurrence, which
should affect the simulation state \code{G} deterministically.
Occurring events are passed as arguments to the callback, which must never
perform any side effects, such as stateful calls or rendering frames.
In our example, we modify the state of the rectangle as follows:
    on \code{START}, reset its position and speed;
    on \code{TICK},  increment its position based on the current speed; and
    on \code{KEY},   modify its speed:

@@

The restriction for stateless calls is the reason why the rendering callback
\code{cb\_ren}~\cite{tml.js} must be separated from \code{cb\_sim}, which would
otherwise render the screen multiple times while time travelling.
Instead, the middleware calls \code{cb\_ren} to render the screen after each
frame event, but only in real time.
%, which the middleware calls only on real-time frames.
In our example, \code{cb\_ren} just needs to redraw the rectangle at the
current position:

@@

The callback \code{cb\_evt} is invoked by the middleware in only real time,
whenever a local SDL event occurs.
This callback has two goals:
    (a) to map and broadcast application events, and
    (b) to provide instant feedback to the user.
%As mentioned previously,
Not all local low-level events need to be broadcast in the network, only those
mapped as application events, as described in Section~\ref{sec.tml.api.events}.
The callback is allowed to perform side effects, such as modifying the network
topology, or exhibiting instant feedback on the screen.
In our example, we only generate application events for the 5 key of interest,
but we also exhibit the pressed key in real time as a visual
feedback:

@@

This section provided an overview of the middleware implementation to answer
how it ensures the desired properties in concrete terms as follows:
    (a) peers meet event deadlines either travelling back and forth or in real
        time (Section \ref{sec.tml.middleware.loop}: \linx{12--16} or
        \linx{18--22});
    (b) peers synchronize their clocks with periodic \code{SYNC} packets that
        fast forward late peers (Section \ref{sec.tml.middleware.loop}:
        \linx{7--10}); and
    (c) peers can join the network at any time and still catch up in time by
        receiving the full event history
        (Section~\ref{sec.tml.middleware.events}).

Two facts ensure consistency in the network:
    (a) peers share a global timeline of events, and
    (b) they go through bit-identical steps
%
However, since the clocks may diverge temporarily, we must rely on simulation
rollbacks.
Therefore, the key to preserve consistency in the presence of rollbacks is to
restrict our programming API to deterministic and stateless calls only.
This way, it becomes impossible to diverge from other peers by, for example,
re-executing stateful file operations multiple times (or at different times).

@@

We run the application main loop at 50 FPS ($20ms$ per frame) and calculate
the middleware overhead at each iteration.
We found a negligible overhead of $9us$ in the average, which corresponds to
less than $1/2000$ of the CPU time.
This means that an application using the middleware would perform all network
synchronization and still have almost 100\% of the CPU time available for its
actual domain logic.
%
In order to have an absolute clock, we use a single machine to simulate all
peers, allowing us to merge their logs into a single file with a comparable
shared timeline.
%
We use \emph{NetEm}~\cite{netem} to simulate the network latency with a normal
distribution.
% (e.g., \code{tc qdisc add dev lo root netem delay 200ms 40ms distribution normal}).
%
We vary
    (i)   the network latency ($5-500ms$ between each peer),
    (ii)  the rate of events ($5-200~evt/sec$), and
    (iii) the deadline $\Delta$ multiplier ($1-5x$).
%
Figure~\ref{fig.mult} shows the $\Delta$s considering the network latency and
chosen multiplier.
For instance, a $50ms$ latency with a $2x$ multiplier results in
$\Delta=500ms$, i.e., each event is triggered with this deadline delay to
compensate the given latency and fixed average of 5 hops
($5 \times 50 \times 2 = 500$).

@@

Our results show that goals (b) and (c) are met in all scenarios consistently,
which are nearly optimal:
Regardless of the network latency, event rate, and delay multiplier,
    the variation in the pace of peers is under $0.2\%$ (goal b), while
    recovering peers take $1.3$ seconds to catch up in time (goal c).
As discussed in Section~\ref{sec.tml}.b, the middleware may take up to $1$
second to catch up, resulting in no more than 2 time travels to recover peers.

Considering goal (a), we need to take a closer look at Figure~\ref{fig.baks}.
The color patterns indicate clear boundaries for the scenarios that can meet
event deadlines with the desired performance.
%
For instance, high event deadlines (column $5x$) make all scenarios viable,
even with high latencies and event frequencies.
As an application example, chats are not affected by higher event deadlines.
Nevertheless, the lower is the deadline, the less applications suit the
middleware.
%
For the lowest possible $\Delta$ (column $1x$), we see that the middleware can
either handle low latencies with high event frequencies (e.g., $5ms$ with
$100evt/min$) or high latencies with low event frequencies (e.g., $100ms$ with
$5evt/min$).
Note that the middleware cannot ensure \emph{instantaneity} for events in most
scenarios, which would require a very low $\Delta$~\cite{melding}.
%
The middleware does not perform well under higher network latencies (e.g., over
$250ms$).
This is expected since peers that are multiple hops away take too long to
synchronize past events and may generate future events in the meantime, which
causes time travels.
%
In summary, the color patterns in the table provide the necessary information
to confront an application with its feasibility or desired performance.

@@

It is expected that centralized solutions exhibit a better performance in
comparison to peer-to-peer applications, since nodes are only 2 hops away from
each other.
%
However, the main benefit of our work is that applications remain functional
even under network partitions, which includes failures in (would be) servers.
Taking this possibility to an extreme, a peer can still work normally even
completely offline for a period of time, as desired by local-first
software~\cite{local}:
Even though rollbacks are inevitable when reconnecting, their effects may be
unnoticeable for final users if there is low activity or dependency between
events.

@@

We propose a middleware to program \emph{symmetric peer-to-peer applications}
in which interacting peers conform to identical behavior without a centralized
coordination server.
%
The middleware API shares similar limitations with centralized
alternatives~\cite{gals,croquet2}, being restricted to deterministic and
stateless calls. %, and only supports pre-allocated memory.
%
%To our knowledge, it is the first middleware to support bit-identical real-time
%behavior in decentralized peer-to-peer networks.
%
Our main contribution is a time machine that supports rollbacks based on memory
snapshots and deterministic simulation.
This allows to decentralize the network timeline, since conflicting peers can
resynchronize their state through time travels.
%
%We assume that peers are non-malicious and that they can assign themselves
%unique identifiers.
%
The middleware provides \emph{perceptive consistency}~\cite{melding}, which
adds a time deadline to reach consistency, ensuring that an event triggered by
any peer at time $t$ is perceived by all peers simultaneously at time
$t+\Delta$.

>>>>

Real-time networked applications allow multiple users to interact remotely and
yet share the same experience.
Examples of these \emph{symmetric distributed applications}~\cite{gals} are
collaborative documents, watch parties, and multi-player games.
%~\cite{TODO:collab}.
%
In order to reproduce the exact behavior in multiple devices, the application
must be able to synchronize time and execution across the network.
In addition, since users can interact asynchronously with the application,
event occurrences must somehow be synchronized back across devices.
%
This problem relates to consistency in distributed systems, but with the
additional requirement of a time deadline to reach consistency~\cite{melding}.

@@

In this work, we propose a middleware for \emph{symmetric peer-to-peer
applications}, which do not rely on central servers for coordination.
Peers in the application form a dynamic network graph and communicate only
with direct neighbours, as in typical unstructured peer-to-peer
networks~\cite{p2p.survey}.
%
Events are flooded in the graph and are applied with a small deadline delay to
compensate the network latency.
To deal with events received out of order or too late, a time machine can
rollback a peer to a previous state and reapply the events in order, such that
a single consistent timeline is established across the network.
%
Our main contribution is the design of a middleware to ensure that all peers
    (a) meet event deadlines,
    (b) advance in sync in real time, and
    (c) can leave and join the network and remain symmetric.
%
Our middleware ensures \emph{perceptive consistency}~\cite{melding}, in which
an event triggered by any peer at time $t$ is perceived by all peers
simultaneously at time $t+\Delta$.
%
As the main limitation, the middleware cannot ensure \emph{instantaneity} for
events in most scenarios, which would require $\Delta$ to be less than a
user-observable threshold~\cite{melding}.
We also assume that peers are non malicious in the sense that they do not lie
about the time in which events occurred locally.

We perform simulations with over 20 peers in a 5-hop mixed topology with cycles
and straight lines.
We vary the network latency and peer churn, as well as the frequency and
deadlines of events.
%
We then measure (a) the recurrence of time travels, (b) the real-time pace of
peers, and (c) the recoverability from churns, such that they validate our
goals above, respectively.
%
As results, we show that goals (b) and (c) are met even under extreme
conditions: peers exhibit a time mismatch of under $0.2\%$ and take less than
$1.5s$ to recover from churns.
%
For the goal (a), we analyze the scenarios in which the middleware can meet
event deadlines with the desired performance.
As an example, peers need to roll back only $3\%$ of the time under a network
latency of 50ms and 25 events per minute.
%
We consider this overhead negligible for soft real-time applications that do
not depend on instantaneous network feedback, such as chats, watch parties, and
turn-based games.

@@

Croquet~\cite{croquet,croquet2,croquet.site} guarantees bit-identical real-time
behavior for users in collaborative distributed environments.
%
As detailed in Figure~\ref{fig.algos}, a centralized \emph{reflector} issues
fixed periodic ticks (e.g., every 50ms), such that all nodes in the network
advance together according to a synchronized global shared clock.
If a user generates an event, it is sent back to the reflector, which
broadcasts the event in the next tick, which all nodes apply in sync.
%
This mechanism is similar to \emph{bucket synchronization}~\cite{gautier}, in
which events are assigned to time buckets to be applied later.
%
Croquet takes periodic snapshots of the whole application state in order to
support late joins to a running session.
The new node just needs to request the latest snapshot and simulate the
remaining events to reach the current running state.
%
As indicated in Figure~\ref{fig.table}, Croquet relies on a centralized
network (a), in which nodes advance in sync (b), and in which event broadcasts
and outcomes depend on the central server to be online (c).

@@

As it is the case for the discussed work in this section, we consider that
middlewares for symmetric applications must comply with \emph{perceptive
consistency}~\cite{melding}, as follows:

@@

An antagonistic approach to deterministic reactions to events is to model the
applications with Conflict-free Replicated Data Types~\cite{crdts}.
CRDTs are designed in such a way that all operations are commutative, so that
the order in which they are applied does not affect the final outcome.
%
As detailed in Figure~\ref{fig.algos}, peers can communicate operations
directly to each other with no central authority.
Also, since operations need not to be ordered, they can be applied at the
very first moment they are generated or received, even if the originating peers
are offline.
%
Note that it is not possible to timestamp commutative operations under a unique
global shared clock.
Hence, there is no notion of a timeline in which peers go through bit-identical
steps, but they do eventually reach the same state~\cite{crdts.eventual}.
%
On the one hand, the main advantage of CRDTs is to support local-first
software~\cite{local}, since they can work offline in the same way as online.
On the other hand, they provide very restrictive APIs (the CRDTs themselves),
and do not support real-time consensus among peers~\cite{crdts.consensus}.
%
As indicated in Figure~\ref{fig.table}, CRDTs work on
    (a) decentralized networks, in which
    (b) peers advance independently, and in which
    (c) operations can be applied immediately, even while offline.
%
Automerge~\cite{p2p.automerge} along with its accompanying Hypermerge
peer-to-peer infrastructure~\cite{p2p.pushpin} is an example of decentralized
CRDTs.

The main advantage of CRDTs is that they can be designed in such a way that
concurrent modifications to shared state can preserve intent~\cite{peritext},
regardless of the order in which they are applied.
%
This is not the case for the symmetric application middlewares described in the
previous section.
Nevertheless, programmers could still use CRDTs on top of a middleware to
preserve intent, since they are nothing more than data structures and merging
algorithms.

Contrasting with perceptive consistency, CRDTs provide \emph{delayed
consistency}~\cite{melding}, as follows:

@@

Regarding local single-user time machines, the video game Braid%
\footnote{Braid: \url{http://braid-game.com/}}
is designed on the assumption that players have unrestricted rollbacks as part
of the game mechanics.
Like Fusion, the implementation is based on continuous memory snapshots,
instead of deterministic simulation as we propose in this work.
%
Schuster~\cite{tml.js} proposes a simple API for time travelling in
event-driven JavaScript:
    \code{init()} to return an initial state;
    \code{handle(evt,mem)} to process events and modify state; and
    \code{render(mem)} to output the current state.
As we describe next, we use a similar approach by separating state mutation
from rendering~\cite{tml.alive}.

@@

Our middleware is written in C and relies on the library \emph{SDL}%
\footnote{SDL: \url{http://www.libsdl.org}}
to deal with the network, timers, and input events, such as mouse clicks and
key presses.
The full source code%
\footnote{Code: \url{http://github.com/fsantanna/p2p-tml/}}
is around 500 lines of code.

@@

In Figure~\ref{fig.middleware}, a peer is depicted as an application using an
API to communicate with the middleware.
The application itself is not aware of the network and has no direct access to
other peers, which communicate transparently through the middleware.
The application generates asynchronous events that go through the middleware,
which synchronizes them with a timestamp with a deadline delay ($\Delta$)
scheduled to the future (aka \emph{local lag}~\cite{mauve}]), such that all
peers can satisfy.
The middleware controls the execution of the application by issuing ticks and
synchronized events.
Note that ticks need not to be broadcast, only $\Delta$ events.

The main job of the middleware is to deal with the uncertainties of the
network in order to provide a consistent timeline and preserve the symmetric
behavior across peers.
As described in the Introduction, the middleware ensures that all peers
    (a) meet event deadlines,
    (b) advance in sync in real time, and
    (c) can leave and join the network and remain symmetric:
%
\begin{enumerate}
\item \textbf{Event Deadlines:}
Due to the inherent latency of networks, a peer (through the middleware)
generates events with a deliberate $\Delta$ so that they can reach other peers
in time to be applied in sync.
%We first assume a predetermined delay for the sake of simplicity.
Nevertheless, a peer ahead of time may receive an event that should have been
applied in the past, which characterizes an optimistic approach to choose
$\Delta$, since it assumes that inconsistencies may occur~\cite{melding}.
%, even considering its $\Delta$.
%
In this case, the middleware automatically rolls back the peer, applies the
event at the correct time, and then fast forwards the peer, re-applying all
remaining events up to the real time.
As we detail further, rolling back a peer requires to simulate the whole
application execution and event occurrences from the beginning.
As an optimization, the middleware takes periodic snapshots locally to avoid
full simulation.
%
\item \textbf{Time Synchronization:}
Since peers run in different machines, their timelines will inevitably be out
of sync because applications are launched locally at different times, and
because internal clocks may diverge over time (e.g., from rollbacks or timer
polling inaccuracies).
The middleware assumes the maximum time among all peers the be the
``correct real time'', so that other peers must fast forward to catch up with
it.
Therefore, in order to determine the maximum time and synchronize the clocks,
the middleware proceeds as follows:
    (a) peers broadcast \code{SYNC} events periodically with their local
        times, and
    (b) if a peer is behind a received \code{SYNC}, it advances its frames
        proportionally faster to catch up in time.
%
To avoid abrupt visual changes, the middleware makes smooth time transitions
that take up to $1s$ to complete.
%
\item \textbf{Peer Churn:}
Considering that we target dynamic peer-to-peer networks, it is important that
the middleware can handle arrival and departure of peers seamlessly.
Note that high churns may lead to intermittent network partitions.
Given our unstructured approach, nothing needs to be done when a peer leaves
the network.
However, when a peer joins the network, the middleware needs to ensure that it
receives from other peer all events that ever happened, in order.
This is trivial since all peers already need to hold the full event history (as
described in item (a)).
% including their absolute timestamps in the shared timeline.
\end{enumerate}

As a limitation of our work, regarding items (a) and (b), we assume that peers
never lie about their local clocks, which would affect event timestamps,
allowing a peer to modify the global timeline with nonsensical information.
%
Therefore, we assume a permissioned peer-to-peer network with an external
mechanism for authentication.

@@

The middleware expects to take full control of the application in order to
manipulate its timeline and disseminate events in the network.
For this reason, the basic programming API is to call a \code{loop} function
passing a set of callbacks, which the middleware uses to communicate with the
application:

@@

Note that using callbacks is already a common technique for event-driven
programming, such as in Games, GUIs, and interactive applications in general,
sharing similarities with \emph{MVC}, \emph{Publish-Subscribe}, and
\emph{Observer} patterns~\cite{meyer,nystrom}.
%
Furthermore, the default run-to-completion semantics of callbacks is key to
ensure the desired determinism, since unlike threads, they run sequentially and
atomically from start to end~\cite{events,threads}.

@@

The middleware calls \code{cb\_sim} every frame or event occurrence, which
should affect the simulation state \code{G} deterministically.
Occurring events are passed as arguments to the callback, which must never
perform any side effects (outside \code{G}), such as stateful calls or
rendering frames.
In our example, we modify the state of the rectangle as follows:
    on \code{START}, reset its position and speed;
    on \code{TICK},  increment its position based on the current speed; and
    on \code{KEY},   modify its speed:

@@

The restriction for stateless calls is the reason why the rendering callback
\code{cb\_ren} must be separated from \code{cb\_sim}~\cite{tml.js}, which would
otherwise render the screen multiple times while time travelling.
Instead, the middleware calls \code{cb\_ren} to render the screen after each
frame event, but only in real time.
%, which the middleware calls only on real-time frames.
In our example, \code{cb\_ren} just needs to redraw the rectangle at the
current position:

@@

The callback \code{cb\_evt} is invoked by the middleware only in real time,
whenever a local SDL event occurs.
This callback has two goals:
    (a) to map and broadcast application events, and
    (b) to provide instant feedback to the user.
%As mentioned previously,
Not all local low-level events need to be broadcast in the network, only those
mapped as application events, as described in Section~\ref{sec.tml.api.events}.
The callback is allowed to perform side effects, such as modifying the network
topology, or exhibiting instant feedback on the screen.
In our example, we only generate application events for the 5 keys of interest,
but we also exhibit the pressed key in real time as a visual
feedback:

@@

This section provided an overview of the middleware implementation to answer
how it ensures the desired properties in concrete terms as follows:
    (a) peers meet event deadlines either in real time or travelling back and
        forth (Section \ref{sec.tml.middleware.loop}: \linx{18--22} or
        \linx{12--16});
    (b) peers synchronize their clocks with periodic \code{SYNC} packets which
        fast forward late peers (Section \ref{sec.tml.middleware.loop}:
        \linx{7--10}); and
    (c) peers can join the network at any time and still catch up in time by
        receiving the full event history
        (Section~\ref{sec.tml.middleware.events}).

Two facts ensure consistency in the network:
    (a) peers share a global timeline of events, and
    (b) they go through bit-identical steps
%
However, since the clocks may diverge temporarily, we must rely on simulation
rollbacks.
Therefore, the key to preserve consistency in the presence of rollbacks is to
restrict our programming API to deterministic and stateless calls only.
This way, it becomes impossible to diverge from other peers by inadvertently
re-executing stateful file operations multiple times (or at different times).

@@

We run the application main loop at 50 FPS (20ms per frame) and calculate
the middleware overhead at each iteration.
We found a negligible overhead of $9us$ in the average, which corresponds to
less than $1/2000$ of the CPU time.
This means that an application using the middleware would perform all
synchronization upkeep and still have almost 100\% of the CPU time available
for its actual domain logic.
%
In order to have an absolute clock, we use a single machine to simulate all
peers, allowing us to merge their logs into a single file with a comparable
shared timeline.
%
We use \emph{NetEm}~\cite{netem} to simulate the network latency with a normal
distribution.
% (e.g., \code{tc qdisc add dev lo root netem delay 200ms 40ms distribution normal}).
%
We vary
    (i)   the network latency ($5-500ms$ between each peer),
    (ii)  the rate of events ($5-200~evt/sec$), and
    (iii) the deadline $\Delta$ multiplier ($1-5x$).
%
Figure~\ref{fig.mult} shows the $\Delta$s considering the network latency and
chosen multiplier.
For instance, a $50ms$ latency with a $2x$ multiplier results in
$\Delta=500ms$, i.e., each event is triggered with this deadline delay to
compensate the given latency and fixed average of 5 hops
($5 \times 50 \times 2 = 500$).

@@

Our results show that goals (b) and (c) are met in all scenarios consistently,
which are nearly optimal:
Regardless of the network latency, event rate, and delay multiplier,
    the variation in the pace of peers is under $0.2\%$ (goal b), while
    recovering peers take $1.3$ seconds to catch up in time (goal c).
As discussed in Section~\ref{sec.tml}.b, the middleware may take up to $1$
second to catch up, resulting in no more than 2 time travels to recover a peer.

Considering goal (a), we need to take a closer look at Figure~\ref{fig.baks}.
The color patterns indicate clear boundaries for the scenarios that can meet
event deadlines with the desired performance.
%
For instance, high event deadlines (column $5x$) make all scenarios viable,
even with high latencies and event frequencies.
As an application example, chats are not affected by higher event deadlines.
Nevertheless, the lower is the deadline, the less applications suit the
middleware.
%
For the lowest possible $\Delta$ (column $1x$), we see that the middleware can
either handle low latencies with high event frequencies (e.g., $5ms$ with
$100evt/min$) or high latencies with low event frequencies (e.g., $100ms$ with
$5evt/min$).
Note that the middleware cannot ensure \emph{instantaneity} for events in most
scenarios, which would require a very low $\Delta$~\cite{melding}.
%
The middleware does not perform well under higher network latencies (e.g., over
250ms).
This is expected since peers that are multiple hops away take too long to
synchronize past events and may generate future events in the meantime, which
results in time travels.
%
In summary, the color patterns in the table provide the necessary information
to confront an application with its feasibility or desired performance.

@@

It is expected that centralized solutions exhibit a better performance in
comparison to peer-to-peer applications, since nodes are only 2 hops away from
each other.
%
However, the main benefit of our work is that applications remain functional
even under network partitions, which includes failures in (would be) servers.
Taking this possibility to an extreme, a peer can still work normally even
while completely offline for a period of time:
%, as desired by local-first software~\cite{local}:
Even though rollbacks are inevitable when reconnecting, their effects may be
unnoticeable to final users if there is low activity or dependency between
events.

@@

We propose a middleware to program \emph{symmetric peer-to-peer applications}
in which interacting peers conform to identical behavior without a centralized
coordination server.
%
The middleware API shares similar limitations with centralized
alternatives~\cite{gals,croquet2}, being restricted to deterministic and
stateless calls. %, and only supports pre-allocated memory.
%
%To our knowledge, it is the first middleware to support bit-identical real-time
%behavior in decentralized peer-to-peer networks.
%
Our main contribution is a time machine that supports rollbacks based on memory
snapshots and deterministic simulation.
This allows us to decentralize the network timeline, since conflicting peers
can resynchronize their state through time travels.
%
%We assume that peers are non-malicious and that they can assign themselves
%unique identifiers.
%
The middleware provides \emph{perceptive consistency}~\cite{melding}, which
adds a deadline to reach consistency, ensuring that an event triggered by any
peer at time $t$ is perceived by all peers simultaneously at time $t+\Delta$.
